### 2017-11-07 submission/manuscript.Rmd
abstract fixes

**Background.** An increasing body of literature suggests that there a role for the microbiota in colorectal cancer. Important players within this axis have ranged from individual microbes to the whole community. A recent meta-analysis investigated this but only focused on potential biomarkers. This study expands on this previous research and tests the hypothesis that detectable changes in the bacterial community are important both to increasing relative risk and model accuracy for adenoma and carcinoma. To test this hypothesis we examined both feces (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) across 14 different studies.

**Results.** There was a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness for fecal samples (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study and variable region. Lower Shannon diversity and evennes in fecal samples resulted in a significant increase in relative risk for carinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community resulted in a higher classification model AUC for both adenoma and carcinoma for fecal and tissue samples. Most studies are adequately powered for large effect size differences which may be more amenable for carcinoma than for adenoma.

**Conclusions.** This data provides support for the idea of the "driver-passenger" model in carinoma but only partially for adenoma. The data is much stronger for carcinoma and this may in part be due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

>>>>

**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer. Important drivers within this axis have ranged from individual microbes to the whole community. A recent meta-analysis investigated whether consistent biomarkers could be identified across studies. This report expands on this previous research and tests the hypothesis that the bacterial community is an important driver of disease from control to adenoma to carcinoma. To test this hypothesis we examined both feces (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) across 14 different studies.

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced. Lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) while only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk. Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community versus only CRC associated genera to build a prediction model resulted in a higher classification AUC for both adenoma and carcinoma for fecal and tissue samples. For the studies that were analyzed, most were adequately powered for large effect size differences which may be more amenable for carcinoma than for adenoma microbiota research.

**Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis. The evidence that has been collated within this study is much stronger for carcinoma and this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

### 2017-11-07 submission/manuscript.tex
abstract fixes

\textbf{Background.} An increasing body of literature suggests that
there a role for the microbiota in colorectal cancer. Important players
within this axis have ranged from individual microbes to the whole
community. A recent meta-analysis investigated this but only focused on
potential biomarkers. This study expands on this previous research and
tests the hypothesis that detectable changes in the bacterial community
are important both to increasing relative risk and model accuracy for
adenoma and carcinoma. To test this hypothesis we examined both feces
(total individuals = 1737) and tissue (total samples = 492) across 14
different studies.

\textbf{Results.} There was a significant decrease from control to
adenoma to carcinoma for both Shannon diversity and evenness for fecal
samples (P-value \textless{} 0.05) after correcting for study and
variable region. Lower Shannon diversity and evennes in fecal samples
resulted in a significant increase in relative risk for carinoma
(P-value \textless{} 0.05) but not adenoma (P-value check tables 0.05).
Previously associated colorectal cancer genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
followed a similar pattern with a significantly increased relative risk
by their presence for carcinoma (P-value \textless{} 0.05) but not
adenoma (P-value \textgreater{} 0.05) with the exception of
\emph{Porphyromonas} (P-value \textless{} 0.05). Using the whole
community resulted in a higher classification model AUC for both adenoma
and carcinoma for fecal and tissue samples. Most studies are adequately
powered for large effect size differences which may be more amenable for
carcinoma than for adenoma.

\textbf{Conclusions.} This data provides support for the idea of the
``driver-passenger'' model in carinoma but only partially for adenoma.
The data is much stronger for carcinoma and this may in part be due to
the low power to detect more subtle changes in the majority of studies
that have been performed to date.

>>>>

\textbf{Background.} An increasing body of literature suggests that
there is a role for the microbiota in colorectal cancer. Important
drivers within this axis have ranged from individual microbes to the
whole community. A recent meta-analysis investigated whether consistent
biomarkers could be identified across studies. This report expands on
this previous research and tests the hypothesis that the bacterial
community is an important driver of disease from control to adenoma to
carcinoma. To test this hypothesis we examined both feces (total
individuals = 1737) and tissue (total samples = 492) across 14 different
studies.

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness (P-value
\textless{} 0.05) after correcting for study effect and variable region
sequenced. Lower Shannon diversity and evenness in fecal samples
resulted in a significant increase in relative risk for carcinoma
(P-value \textless{} 0.05) while only evenness for adenoma (P-value
\textless{} 0.05) resulted in a slightly increased relative risk.
Previously associated colorectal cancer genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
followed a similar pattern with a significantly increased relative risk
by their presence for carcinoma (P-value \textless{} 0.05) but not
adenoma (P-value \textgreater{} 0.05) with the exception of
\emph{Porphyromonas} (P-value \textless{} 0.05). Using the whole
community versus only CRC associated genera to build a prediction model
resulted in a higher classification AUC for both adenoma and carcinoma
for fecal and tissue samples. For the studies that were analyzed, most
were adequately powered for large effect size differences which may be
more amenable for carcinoma than for adenoma microbiota research.

\textbf{Conclusions.} This data provides support for the importance of
the bacteral community to both adenoma and carcinoma genesis. The
evidence that has been collated within this study is much stronger for
carcinoma and this may be in part due to the low power to detect more
subtle changes in the majority of studies that have been performed to
date.
